# Introduction
This module is about creating visualizations in R, using the **ggplot2** package.  We will add to the basic information about visualization covered in the pre-course work but continue to focus on bar charts, line charts, and histograms.

- **ggplot2** is among the most useful and widely used for visualization in R.  

- Plots are constructed in layers, which can be more intuitive.

- This package is among the most useful and widely used for visualization in R, created by Hadley Wickham.  

- There are plotting functions available in base R but **ggplot2** provides a powerful, intuitive framework for creating and customizing visualizations.

Here's some examples:
<http://www.r-graph-gallery.com/portfolio/ggplot2-package>


## ggplot2 Package

The **ggplot2** package is based on the grammar of graphics, which breaks graphics into parts which are controlled separately and combined.  Every graph has the same basic components - data, a coordinate system, and geoms that represent data points visually.  

First, lets install the packages we will use in this module.
```{r warning=FALSE,comment=FALSE,message=FALSE}
library(ggplot2)
library(RColorBrewer)
library(plyr)
library(dplyr)
library(scales)
```

Also, clear your Global Environment.
```{r}
rm(list = ls())
```


## References

Some useful references include:

- <http://www.cookbook-r.com>

- <http://ggplot2.tidyverse.org/reference/>

- <https://www.rstudio.com/resources/cheatsheets/>

- "R Graphics Cookbook" by Winston Chang  (PDF can be found online at <http://ase.tufts.edu/bugs/guide/assets/R%20Graphics%20Cookbook.pdf>)

- <http://www.google.com>


## Layering
Every graph has the same basic components: 

1.  Data
* usually a data frame 

2.  Aesthetic mapping
* how variables in the data frame map to visual objects on the graph
* objects (or aesthetics) to map to include x-axis, y-axis, size, shape, color, transparency
* specified with the `aes()` function

3.  Geoms that represent data points visually.  
* specifies the type of plot
* geom = geometric object
* examples (see the cheatsheet!): `geom_line()`, `geom_point()`, `geom_bar()`, `geom_histogram()`  
