# Bar Charts

Bar charts are generally used to display numeric values on the y-axis for different categories or discrete values on the x-axis.  Bar charts are not well suited for continuous data on the x-axis because try to make a bar for every possible value of the continuous range.  

The heights of the bar on the y-axis can represent counts or values from your data set.  The code for the graphic changes depending on which is the case. 


## GDELT Data
For the first part of this module we will use the data from the earlier modules from GDELT, <http://www.gdeltproject.com>.  The dataframe is events in Nigeria that involve Chinese actors in January 2017.

If you don't already have it in your environment, open the "subsahara_jan17.csv" file.

```{r }
subsahara_jan17 <- read.csv("subsahara_jan17.csv")
```


Lets explore the QuadClass variable, which indicates one of four types of events.

1. Verbal Cooperation

2. Material Cooperation

3. Verbal Conflict

4. Material Conflict
```{r}
summary(subsahara_jan17$QuadClass)
```

## Charts with Counts
Let's create a bar chart the shows the number of events for each level of the QuadClass variable.  First, we start with the *ggplot* function to define the data and basic aesthetics, in this case which variable we want to use.  Aesthetics can also include visual properties like size and shape, which we will cover later.
```{r }
ggplot(data = subsahara_jan17, aes(x = QuadClass))

```

Next we need to add a geom function, in this case *geom_bar()*.  This determines the basic graphical look.
```{r echo = FALSE}
library(ggplot2)
```

```{r}
ggplot(data = subsahara_jan17, aes(x = QuadClass)) +
  geom_bar()

```

