# Change the Look
Now we want to adjust the look of the chart by applying more aesthetic controls inside the geom function and using themes and scales.  

We will change:

- Color of the bars

- Width of the bars.

- Axis labels.

- Chart title.



## Colors
- R has over 600 colors built in which you can refer to by name.  

- The *colors()* function provides a list of the names.  

- You can also select colors by hexadecimal codes, such as "#990000" for deep red.  

- [R Cookbook Graphs section](http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/) has a very useful section on colors and palettes.

- Another good reference for color palettes is <http://colorbrewer2.org>

- The *RColorBrewer* package provides a set of useful color palettes that can be used to customize the color ranges on a graph.  Use *display.brewer.all()* to see the set of palettes. 

- Note:  Sometimes you will see the word "color" spelled "colour" - they are interchangeable in R.

## Changing Bar Colors

- Change the fill color of the bars with `fill = "lightblue"`

- Change the color of the outline of the bars with `color = "darkblue"`

- Change with width of the bars with `width = .5`
```{r echo = FALSE}
library(ggplot2)
subsahara_jan17 <- read.csv("subsahara_jan17.csv")
```

```{r}
ggplot(subsahara_jan17, aes(x = QuadClass)) +
  geom_bar(fill = "lightblue", color = "darkblue", width = .5)
```




## Themes
There are many more options for customizing a graph.  Adjusting the look involves of two different types of elements, themes and scales.

- Theme elements are non-data elements in the plot such as the title, legend, and axes.  

- To change the appearance of a theme element, use the *theme()* function and a corresponding element object.  

- Common element objects are text or lines.  

- [R Graphics Cookbook](http://ase.tufts.edu/bugs/guide/assets/R%20Graphics%20Cookbook.pdf) covers this in section 9.4 very clearly. 

Here's an example, changing the color of the y-axis title to red:

```{r eval = FALSE}
theme(axis.title.y = element_text(color = "red"))
```

A few other useful theme functions:

- Remove an axis label: `theme(axis.title.x = element_blank())`

- Remove grid lines: `theme(panel.grid.major = element_blank())`

- Rotate x-axis text labels by 90 degrees:  `theme(axis.text.x = element_text(angle = 90))`

- Move the position of a legend: `theme(legend.position = "bottom")`

## Scales
Scale elements map data values to visual values.  

- To change the default mapping, a new scale is added.  

- The scale function identifies the aesthetic you want to adjust, a pre-packaged scale to use, and the specific arguments for what you are trying to change.  

- Different types of data use different types of aesthetics and scales - it is important to match them up correctly.

- The [Data Visualization Cheatsheet](https://www.rstudio.com/resources/cheatsheets) are a good reference for different types of scales.

Here's an example, using `scale_x_reverse()`:

- "scale" indicates that we want to customize a data element.  

- "x" indicates which data element.

- "reverse" is used to override the default linear mapping and reverse the order on the x-axis.

```{r}
ggplot(subsahara_jan17, aes(x = QuadClass)) +
  geom_bar(fill = "lightblue",color = "darkblue", width = .5)+
  scale_x_reverse()

```



## Labels and Titles
Going back to our bar chart of the Quad Class varaible, next we will adjust the labels and add a chart title.

- Change the label on the x-axis with `xlab("Quad Class of Events")`

- Delete the label on the y-axis (currently "count") with `theme(axis.title.y = element_blank())`

- Add a chart title with `ggtitle("Number of Events by Quad Class")`

```{r}
ggplot(subsahara_jan17, aes(x = QuadClass)) +
  geom_bar(fill = "lightblue", colour = "darkblue", width = .5) +
  xlab("Quad Class of the Events") +
  theme(axis.title.y = element_blank()) +
  ggtitle("Number of Events by Quad Class")
```

## Change the Look Exercise
1.  Change the color of the x-axis label.  (hint: use a theme function)
2.  Increase the size of the chart title. (hint:  use a theme function)
3.  Change the labels on the x-axis to One, Two,Three, and Four.  (hint:  `scale_x_continuous()`)
4.  Change the y-axis label to "Number of Events".

```{r echo = FALSE}
ggplot(subsahara_jan17, aes(x = QuadClass)) +
  geom_bar(fill = "lightblue", colour = "darkblue", width = .5) +
  xlab("Quad Class of the Events") +
  ylab("Number of Events") +
  ggtitle("Number of Events by Quad Class") +
  theme(axis.title.x = element_text(color = 'blue')) +
  theme(plot.title = element_text(size = 20)) +
  scale_x_continuous(breaks = c(1,2,3,4), labels = c("One","Two","Three","Four"))
```

