# Line Charts
Line charts are typically used when all variable are continuous.  Often, line charts are better for graphics involving dates and/or times (often called time series charts) because trends are more visible.


```{r echo = FALSE, warning=FALSE, comment=FALSE}
library(dplyr)
library(ggplot2)
library(RColorBrewer)
china_2weeks <- read.csv("china_2weeks.csv", stringsAsFactors = FALSE)
dailybreaks <- seq(min(as.Date(china_2weeks$days)),
                   max(as.Date(china_2weeks$days)),
                   by="2 day")
```

Let's create some simple time series data with the "china_2weeks" dataframe.
```{r}
china_all <- summarize(group_by(china_2weeks, days), total = sum(n))
china_all$days <- as.Date(china_all$days)

```

Notice that there are no dates that are skipped - we have data for every day.  That's great because it means we don't need to fill in the gaps.

## Basic Line Charts
In **ggplot**, line charts use geom function *geom_line()*.

Let's create a plot of the total events by day.

```{r}
ggplot(china_all, aes(x = days, y = total))+
  geom_line()
```


## Multi-Variable Line Charts
You can also plot multiple lines on a chart.  We will use the china_2weeks dataframe and plot a line for each country.

- To change the color of the line based on country, we need to change the `fill` aesthetic to `color`.  

- You must have another aesthetic besides x and y so that *ggplot* knows how to group the data.  

```{r}
ggplot(china_2weeks, aes(x = as.Date(days), y = n, color = country))+
  geom_line()
```


## Adjust the Look
We can make some adjustments to the look of the line chart.

A good reference for points and lines: <http://www.cookbook-r.com/Graphs/Shapes_and_line_types/>

- Map the country to line type instead of color with `linetype = country`

- Increase the width of the line with `geom_line(size = 1)`

- Add points and change the shape and size of the point with `geom_point(shape = 15, size = 2)`

```{r}
ggplot(china_2weeks, aes(x = as.Date(days), y = n, linetype = country))+
  geom_line(size = 1) +
  geom_point(shape = 15, size = 2)
```


