# Histograms
A histogram is used to map a continuous variable to the x-axis and use bins to depict the distribution of that variable.

- In **ggplot**, histograms use geom function *geom_histogram()*.

- In the subsahara_jan17 dataset, a continuous variable is NumMentions, which is the number of mentions of the event across all source documents, an indicator of the importance of the event.

- We want to see the distribution of the NumMentions variable.

```{r echo = FALSE}
library(ggplot2)
subsahara_jan17 <- read.csv("subsahara_jan17.csv")
```

```{r warning = FALSE, comment=FALSE}
ggplot(subsahara_jan17, aes(x=NumMentions))+
  geom_histogram()
```

The default in **ggplot** is 30 bins.  But we can set the number of bins in two ways.
First we want to find the max and min of the variable.
```{r}
summary(subsahara_jan17$NumMentions)
```

To set the width of the bins use `binwidth = 2`
This will create bins from 1 up to but not including 3, 3 up to but not including 5, and so on.

```{r}
ggplot(subsahara_jan17, aes(x = NumMentions)) +
  geom_histogram(binwidth = 2, fill = "lightblue", color = "black")
```

To set the number of bins use `bins = 5`
```{r}
ggplot(subsahara_jan17, aes(x = NumMentions)) +
  geom_histogram(bins =5, fill = "lightblue", color = "black")
```

